Class imbalance is overwhelmingly handled by oversampling, yet SMOTE-family methods
typically trade majority-class accuracy for minority-class recall, and it is rarely
clear \emph{why} oversampling helps. We introduce a \emph{remedy
decomposition} of the misclassified minority: combining an aleatoric--epistemic
uncertainty decomposition with a threshold-recoverability test, each minority error
is assigned to the remedy that would fix it: \emph{threshold-recoverable} (the
model already ranks it correctly; only the default cutoff is wrong),
\emph{data-reducible} (more same-class data would help), or \emph{irreducible}
(genuine class overlap). Across 79 real datasets the deficit is overwhelmingly
threshold-recoverable (mean $69\%$; $74\%$ at imbalance ratio $>3$); only
$\sim$7\% is data-reducible and $\sim$16\% irreducible. This explains a known but
unexplained regularity: SMOTE-family resampling does not improve ranking (AUC is
unchanged) and a threshold move reproduces its benefit without synthetic
generation; single undertrained neural networks, whose ranking imbalance itself
corrupts, are the exception. Because the deficit
is mostly a thresholding artefact,
such oversampling is at best an implicit, inefficient enactment of a threshold
shift. We give the
mechanism. On controlled mixtures with a \emph{known} Bayes boundary, standard SMOTE
places an overlap-dependent fraction of synthetic points \emph{across} it; a short
proposition shows why interpolating boundary seeds crosses into majority-posterior
space. Empirically, once every strategy receives the same out-of-fold threshold
tuning, the large default-threshold gap between non-generative and generative
families collapses to within $\pm1$\,pp across three base learners (against a
$+6.4$\,pp effect from tuning the threshold alone), so the SMOTE family's apparent
edge is an operating-point effect, not a discrimination gain. The practical conclusion:
diagnose the deficit; for most of it, move the decision rather than oversample.
