\section{Discussion}\label{sec:discussion}

\paragraph{The tradeoff is a decision problem, not a data problem.} Our central
finding is that the minority-class deficit \emph{decomposes by remedy} and is
predominantly \emph{threshold-recoverable}: across 79 datasets a mean of $68\%$ of
minority errors ($80\%$ at $\mathrm{IR}>3$) are corrected by moving the decision
threshold alone, with only $\sim$6\% data-reducible and $\sim$17\% irreducible.
This explains a regularity the field has documented but not explained, namely that
resampling leaves ranking (AUC) unchanged and a threshold shift reproduces its
benefit without synthetic generation and without shifting the probability scale
\cite{provost2000imbalanced,elor2022smote,goorbergh2022harm,dalpozzolo2015calibrating},
and reframes the long-observed accuracy/recall tradeoff: oversampling is an
implicit, inefficient enactment of a threshold move (it shifts the operating point
by generating across the overlap boundary, at an accuracy and calibration cost),
not a way to build a better classifier. Consistent with this, once every strategy
is given the same threshold tuning the large default-threshold gap between the
families collapses to within $\pm1$\,pp.

\paragraph{Exclusion vs.\ targeting.} The dominant paradigm in oversampling
research \emph{targets} sparse or boundary regions for generation
(ADASYN \cite{he2008adasyn}, Borderline-SMOTE \cite{han2005borderline}, and a long
tail of variants \cite{kovacs2019smote}). We acknowledge their design intent:
boundary-targeting variants are often deployed deliberately to buy true-positive
rate at some true-negative cost, in applications where missed positives dominate
the loss. Our results do not say that operating point is wrong; they say a tuned
threshold reaches \emph{any} such recall-biased operating point explicitly,
with calibrated probabilities, and without synthetic data
(\S\ref{sec:results:frontier}). Our mechanism explains why boundary targeting has
not resolved the accuracy tradeoff itself: these variants all generate at or near
the boundary, which
is exactly where generation manufactures majority-territory ``minority'' points.
Consistent with this, masking the boundary seeds from SMOTE removes both the
accuracy cost \emph{and} the balanced-accuracy benefit (corroborating that the
benefit \emph{is} the boundary generation), so the productive move is not a
better oversampler but a different lever entirely.

\paragraph{The menu matters more than the selector.} A recent line of work asks
which data characteristics should pick the resampler
\cite{jiang2026beyond,carvalho2025resampling}, but its rules barely beat
always-SMOTE. We find the reason: given a menu that includes the dominant
non-generative strategies, even an imbalance-ratio-only rule beats always-SMOTE,
and a learned selector over rich error-structure and complexity meta-features adds
only a modest, robust increment over the single best fixed strategy (balanced
random forest), significant on balanced accuracy and G-mean, absent on MCC. The
practical recommendation is therefore not a sophisticated selector but a change of
default: prefer a strong non-generative strategy, and reserve oversampling for the
$\sim$15\% of regimes where the map (\S\ref{sec:results:regime}) says it
wins. Class-weighted learning is an equally legitimate non-generative default
where it suits the pipeline, with one caveat our parity analysis quantifies
(\S\ref{sec:results:frontier}): how completely weighting enacts the
operating-point shift is learner-dependent (fully for logistic regression,
partially for boosting, barely for bagged forests), so explicit threshold-moving
remains the more uniformly reliable handle.

\paragraph{Relation to prior categorization and complexity work.} Napiera{\l}a
\& Stefanowski's instance types \cite{napierala2016types} and data-complexity
measures \cite{komorniczak2022data,santos2023unifying} characterize imbalanced
problems descriptively, and meta-learning relates those characteristics to
observed performance \cite{carvalho2025resampling}. Our contribution is
orthogonal and mechanism-level: the aleatoric--epistemic decomposition localizes
the boundary/overlap region, and the interventional design probes \emph{why}
boundary generation hurts, rather than reporting a correlation. This
intervention-supported account is what the otherwise extensive selection
literature lacks.

\paragraph{Toward deep imbalanced learning.} Our base learners are tabular;
the analysis is
deliberately a controlled, mechanism-level foundation laid before the deep case,
and the six-learner parity sweep already contains the first empirical bridge: a
single early-stopped MLP is the one learner whose \emph{ranking} imbalance
corrupts (AUC $0.769$, restored to $0.892$ by resampling,
\S\ref{sec:results:frontier}), so for undertrained neural learners resampling
has genuine representation value beyond the operating point, while bagging the
same MLPs restores the ranking and with it the threshold-recoverable
decomposition (\appref{app:nontree}). Two
of the account's three pillars are architecture-agnostic, and we expect them to carry over. The
remedy decomposition needs only a ranking score and an aleatoric--epistemic signal,
both available from deep models via deep ensembles or MC-dropout
\cite{lakshminarayanan2017simple}; and the threshold-recoverability result is the same
operating-point logic that already explains why imbalanced-trained neural networks are
systematically miscalibrated and recover under post-hoc threshold/temperature
adjustment \cite{guo2017calibration}. The boundary-generation mechanism is the part
that needs care: deep oversampling typically interpolates in a \emph{learned latent}
space rather than the input space (e.g.\ DeepSMOTE \cite{dablain2022deepsmote}), so
Proposition~\ref{prop:cross} would apply to the \emph{latent} posterior, whose geometry
the encoder itself shapes. We therefore present these results as the necessary tabular
proof-of-mechanism and offer a concrete, testable hypothesis: latent-space
interpolation across the embedded decision boundary should reproduce the same
accuracy/recall tradeoff, and a latent threshold move should likewise recover most of
the minority deficit. Testing it across architectures is the natural next thread.

\paragraph{Limitations.} (i) The learned selector's edge over a strong fixed
default is small ($\sim$0.6\,pp) and metric-dependent (present on balanced
accuracy and G-mean, absent on MCC); we report it honestly rather than as a
headline, and conclude that a fixed non-generative default is the pragmatic
recommendation. (ii) The triage thresholds ($\tau_{\text{noise}}=0.12$,
$\tau_{\text{cat2}}=0.4$) are fixed across datasets; a sensitivity analysis
(\appref{app:sensitivity}) shows category assignments and downstream
accuracy are stable across a wide range. (iii) We validate the default-threshold
advantage (and its collapse to within $\pm1$\,pp at threshold parity) across
six base learners (gradient-boosted trees, random forests, logistic regression,
naive Bayes, RBF-SVM, and MLP; \S\ref{sec:results:frontier}), so neither the
default-threshold gap nor
the operating-point explanation is a model-specific artefact; the single MLP is
the documented exception (its ranking, not just its threshold, degrades under
imbalance), which delimits the account's scope and motivates the deep extension
sketched above. (iv) The regime map is descriptive: a depth-3 rule does not beat
``always use a balanced ensemble,'' so the map informs rather than automates the
choice. (v) The triage's geometry signals (the local class ratio and the
neighbour-based noise signal) use Euclidean $k$-NN on the loaded feature matrix,
with categoricals ordinal-encoded and no standardization; the forest-derived
signals are scale-invariant, but the $k$-NN signals are not, so the
Cat\,2/Cat\,3 boundary can in principle depend on feature units. The threshold
sweep of \appref{app:sensitivity} covers part of this axis
($\tau_{\text{cat2}}\in[0.2,0.6]$ leaves downstream conclusions stable), but a
fully unit-invariant geometry (standardized or mixed-type distances) is a
worthwhile robustness extension.
