\appendix

\section{Threshold Sensitivity}\label{app:sensitivity}

The triage thresholds ($\tau_{\text{noise}} = 0.12$, $\tau_{\text{cat2}} = 0.4$)
and the triage-weighting upweight factor ($w = 2.0$, the sample-weight multiplier
that the triage-weighting strategy of \S\ref{sec:method:menu} applies to the correct
same-class neighbours of Cat\,2 errors) are fixed across all datasets. We swept each threshold across a
range of values and measured downstream XGBoost accuracy: $\tau_{\text{noise}}
\in [0.05, 0.30]$, $\tau_{\text{cat2}} \in [0.2, 0.6]$, $w \in [1.0, 5.0]$.
Across the benchmark datasets, 74\% show $< 0.5\%$ accuracy
variation across all configurations, supporting the use of fixed thresholds
as a sensible default. The two methods (clean-masked SMOTE, triage
weighting) maintain their relative ordering vs.\ baselines at every threshold
configuration tested.

\section{Out-of-Fold Triage Validation}\label{app:oof}

The triage categories underpin the mechanistic argument (\S\ref{sec:results:interventional}),
so we verify they are not optimistic in-sample artefacts. Two facts establish this.
First, by construction every model-derived signal that drives the categorization is
\emph{out-of-bag}: the error indicator is the OOB ensemble majority vote, and the
true-class probability is the OOB estimate (\S\ref{sec:method:uncert}); the only
non-model signal, the local class ratio, is a property of the data geometry.
Second, as a strict check we re-derive the categories under inner $5$-fold
cross-validation: forests fit on the inner-training folds categorize the held-out
instances (via held-out predictions for the error indicator, true-class
probability, and the full aleatoric/epistemic decomposition; neighbour-based
signals are measured against the inner-training set). Over the 79 datasets, the
Cat\,3 (irreducible) fraction among errors is essentially unchanged: the in-sample
and out-of-fold means coincide at $0.838$, the per-dataset Spearman correlation is
$0.97$ ($p<10^{-40}$), and the mean absolute shift is $\OOFCATSHIFT$\,pp (median
$1.0$\,pp). The aleatoric/epistemic ordering that motivates the categories is
preserved out-of-fold (Cat\,3 aleatoric uncertainty exceeds Cat\,2's on $80\%$ of
datasets). The category \emph{identification} is therefore not an in-sample
artefact. We note separately that the small real-data \emph{interventional} effect
(\S\ref{sec:results:interventional}) is sensitive to this re-estimation: with
out-of-fold seeds and a reduced $3\times3$ protocol the targeted-augmentation
contrast remains directionally consistent (random augmentation null) but is no
longer individually significant, which is why we rest the mechanism on the
controlled experiment (\S\ref{sec:results:mechanism}) and the
discrimination/threshold analysis (\S\ref{sec:results:frontier}) rather than on the
real-data intervention.

\section{GBM Triage Agreement}\label{app:gbm}

To verify that the three-way decomposition is not specific to random
forests, we replaced the RF ensemble with bagged XGBoost (5 bags of 100
trees, matching the RF configuration). On a 20-dataset agreement subset,
the GBM-based triage assigns 82\% of training instances to the same category
as the RF-based triage. Downstream accuracy under both triages is
comparable (mean absolute difference $0.04\%$), indicating that any
ensemble model exposing per-tree predictions can drive the decomposition.

\section{Subsampling Category-Structure Control}\label{app:subsampling}

We verified that stratified subsampling to 10\,000 instances preserves the
triage category distribution. On the five largest datasets in our
benchmark, we computed the per-category fraction $\{p_{\text{correct}},
p_{\text{noise}}, p_{\text{data-limited}}, p_{\text{irreducible}}\}$ on the
full dataset and on a single stratified 10\,000-instance subsample. The
mean absolute deviation between the two distributions is $< 2$ percentage
points per category across all five datasets, validating the subsampling
protocol used throughout (\S\ref{sec:results:frontier}).
Detailed per-dataset numbers are provided in the supplementary material.

\section{Computational Cost Detail}\label{app:cost}

Full cost-table breakdown by dataset (extending \S\ref{sec:results:cost}):

\input{tables/table7_cost.tex}

The triage time scales roughly as $O(n \log n)$ in the number of training
instances and is dominated by the random-forest fits; the local class-ratio
computation (k-NN with $k=5$) contributes $< 5\%$ of total triage time
across all dataset sizes.

\section{Instrument Robustness: Non-Tree Ensembles}\label{app:nontree}

The remedy decomposition's model-derived signals come from a random-forest
ensemble (\S\ref{sec:method:uncert}). To verify the headline is not an
artefact of tree-based uncertainty, we re-derive the full decomposition with
two bagged non-tree instruments over the full 27 KEEL + 54
OpenML/scikit-learn roster (81 dataset cells; the headline statistics of
Table~\ref{tab:reducibility} use the 79-dataset menu roster of \S\ref{sec:design}): $B{=}25$
bootstrap-bagged multilayer perceptrons (scaled, $64{\times}32$ hidden units,
early stopping) and $B{=}25$ bootstrap-bagged logistic regressions. Every
signal remains out-of-bag: only ensemble members whose bootstrap excluded an
instance contribute to its probabilities, and the Shaker-style
aleatoric--epistemic decomposition is taken over those members. The local
class ratio is the identical $k$-NN computation, and the error set, one-vs-rest
intervention, and cross-fitted threshold follow \S\ref{sec:method:remedy}
exactly. Two design notes: (i)~members
are unweighted, matching the headline instrument (\S\ref{sec:method:remedy}) ---
class-balanced members enact the operating-point shift
\emph{inside} the instrument (completely so for logistic regression,
\S\ref{sec:results:frontier}) and dissolve the default-threshold deficit being
decomposed; (ii)~Cat\,1 is gated by TCP alone, as the forest-consensus noise
score has no non-tree analogue, which shifts part of the residue from the
data-reducible/irreducible classes into noise but leaves the
threshold-recoverability axis untouched.

\begin{table}[t]
\centering
\caption{Remedy decomposition by instrument (means over datasets with
$\ge 5$ minority errors; definitions of \S\ref{sec:method:remedy}). The
threshold-recoverable share, the paper's headline, is stable across
instruments; the per-dataset correlation with the RF instrument is given in
the last column. The non-tree instruments assign more of the residue to noise
(TCP-only gate, note (ii)).}
\label{tab:nontree}
\footnotesize
\input{tables/table_nontree.tex}
\end{table}

Threshold-recoverable remains the dominant remedy class under both non-tree
instruments (Table~\ref{tab:nontree}): $62\%$ for bagged MLPs and $63\%$ for
bagged logistic regression against $68\%$ for the RF instrument, rising to
$73\%$ on $\mathrm{IR}>3$ datasets for both (RF: $80\%$), with per-dataset
correlations of $r=0.78$ and $0.72$ against the forest instrument. Part of the
gap to the RF share is the TCP-only noise gate of note (ii), which absorbs
$19$--$24\%$ of the residue into noise. The decomposition's ordering,
threshold-recoverable far above every other remedy class, is
instrument-independent.

\section{Ensemble-Size Sensitivity}\label{app:msens}

The instrument uses $M{=}5$ forests of $T{=}100$ trees, matching standard
deep-ensemble practice ($M{=}5$ in \cite{lakshminarayanan2017simple}).
Recomputing the full decomposition for $M \in \{1,2,5,10,20\}$ on a
10-dataset IR-spanning subset (Table~\ref{tab:msens}): the triage category
fractions move by at most $2.6$\,pp between $M{=}5$ and $M{=}20$, and the
threshold-recoverable share by $3.3$\,pp on average ($1.9$\,pp for $M{=}10$);
the largest single-dataset deviations occur where the minority error set is
tiny (a handful of instances, so one flip moves the fraction by many points).
Below $M{=}5$ the decomposition is visibly unstable (single-dataset swings up
to $34$\,pp). $M{=}5$ therefore sits at
the stability knee, and the conclusions do not change with a larger ensemble.

\begin{table}[t]
\centering
\caption{Ensemble-size sensitivity: mean threshold-recoverable share and
absolute deviations from the $M{=}5$ reference across 10 datasets.}
\label{tab:msens}
\footnotesize
\input{tables/table_msens.tex}
\end{table}

\section{Extended Oversampler Comparison}\label{app:oversamplers}

Table~\ref{tab:smote} reports the full oversampler family, including the
Kov\'acs top-ranked variants and Geometric SMOTE (added at review), against
standard SMOTE on the OpenML roster.

\input{tables/table3_smote.tex}

\clearpage
\section{Dataset Listing}\label{app:datasets}

Table~\ref{tab:datasets} lists the OpenML/scikit-learn roster, including
each dataset's OpenML ID, number of instances, features, classes, and imbalance
ratio (majority-class count divided by minority-class count); the remaining 27
datasets are the standard \texttt{imbalanced-learn}/KEEL benchmark suite.

\input{tables/table_datasets.tex}
